---
title: "Cat Project Analysis"
author: "Group-A: Josh and Olivia"
date: "2/19/25"
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r, setup, include=FALSE}
require(dplyr)
require(mosaic)   # Load additional packages here 
require(Amelia)
library(ggplot2)
library(tidyr)
library(kableExtra)

# Some customization.  You can alter or delete as desired (if you know what you are doing).
trellis.par.set(theme=theme.mosaic()) # change default color scheme for lattice
knitr::opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
```

```{r include=FALSE}
cats <- read.csv("http://pmatheson.people.amherst.edu/CatsSet.csv")
```

``` {r include=FALSE}
#Get rid of pesky NA's in quantitative sets and replace with the average
cats$Age[is.na(cats$Age)] <- mean(cats$Age,na.rm=T)
cats$Weight..lbs.[is.na(cats$Weight..lbs.)] <- mean(cats$Weight..lbs.,na.rm=T)
cats$pH[is.na(cats$pH)] <- mean(cats$pH,na.rm=T)
cats$BG[is.na(cats$BG)] <- mean(cats$BG,na.rm=T)
cats$iCa[is.na(cats$iCa)] <- mean(cats$iCa,na.rm=T)
cats$K[is.na(cats$K)] <- mean(cats$K,na.rm=T)
cats$Lac[is.na(cats$Lac)] <- mean(cats$Lac,na.rm=T)
cats$BUN[is.na(cats$BUN)] <- mean(cats$BUN,na.rm=T)
cats$Creat[is.na(cats$Creat)] <- mean(cats$Creat,na.rm=T)
cats$TS[(cats$TS)=="No bloodwork"] <- NA
cats$PCV[(cats$PCV)=="No bloodwork"] <- NA
cats$TS[(cats$TS)==""] <- NA
cats$PCV[(cats$PCV)==""] <- NA
cats$PCV<-as.numeric(cats$PCV)
cats$TS<-as.numeric(cats$TS)
cats$PCV[is.na(cats$PCV)] <- mean(cats$PCV,na.rm=T)
cats$TS[is.na(cats$TS)] <- mean(cats$TS,na.rm=T)
cats$Temp[is.na(cats$Temp)] <- mean(cats$Temp,na.rm=T)
cats$HR[is.na(cats$HR)] <- mean(cats$HR,na.rm=T)
cats$Urine.protein[(cats$Urine.protein)=="Not checked"] <- NA
cats$Urine.protein<-as.numeric(cats$Urine.protein)
cats$Urine.protein[is.na(cats$Urine.protein)] <- mean(cats$Urine.protein,na.rm=T)
cats$USG[(cats$USG)=="Not checked"] <- NA
cats$USG<-as.numeric(cats$USG)
cats$USG[is.na(cats$USG)] <- mean(cats$USG,na.rm=T)
cats$Urine.pH[(cats$Urine.protein)=="Not checked"] <- NA
cats$Urine.pH<-as.numeric(cats$Urine.pH)
cats$Urine.pH[is.na(cats$Urine.pH)] <- mean(cats$Urine.pH,na.rm=T)

cats$Diet[(cats$Diet=="DRY")] <- "Dry"
cats$Diet[(cats$Diet=="Unknown")] <- "Dry"
cats$Access.to.outdoors.[(cats$Access.to.outdoors.=="Unknonw")] <- "No"
cats$Access.to.outdoors.[(cats$Access.to.outdoors.=="Unknown")] <- "No"
cats$Stress[(cats$Stress=="Unknown")] <- "No"
cats$Previous.FLUTD.signs.[(cats$Previous.FLUTD.signs.=="Unknown")] <- "Yes"
cats$Vx.[(cats$Vx.=="Unknown")] <- "Yes"
cats$PMHx[(cats$PMHx=="Unknown")] <- "No"

cats$Size.of.u.cath[(cats$Size.of.u.cath=="Slippery Sam")] <- 3.5
cats$Flush.[(cats$Flush.=="Unknown")] <- "Yes"
cats$Bolus.[(cats$bolus=="Unknown")] <- "Yes"
cats$Severe.illness.[(cats$Severe.illness.=="Undetermined")] <- "No"

cats$Anemia.[(cats$Anemia.=="Not checked")] <- "No"
cats$Became.anemic.[(cats$Became.anemic.=="Not rechecked")] <- "No"
cats$Became.anemic.[(cats$Became.anemic.=="No bloodwork")] <- "No"
cats$Crystals[(cats$Crystals=="Not checked")] <- "Yes"
cats$Urine.WBC[(cats$Urine.WBC=="Not checked")] <- NA
cats$Urine.RBC[(cats$Urine.RBC=="Not checked")] <- NA


data2<-na.omit(cats) #work with this data - doesn't have any N/As

```

```{r include = FALSE}
CatDatasetCreat <- cats %>%
  mutate(Ket = ifelse(Ketamine== "Yes", 1, 0)) %>%
  mutate(Val = ifelse(Valium=="Yes", 1, 0)) %>%
  mutate(Opi = ifelse(Opioid== "Yes", 1, 0)) %>%
  mutate(Dex = ifelse(Dexmedetomidine=="Yes", 1, 0)) %>%
  mutate(Iso = ifelse(Isoflurane=="Yes", 1, 0)) %>%
  mutate(Prop = ifelse(Propofol=="Yes", 1, 0)) %>%
  rename(RepeatUO="Repeat.UO.")%>%
  rename(Survival30="Survival.30.day")%>%
  rename(FLUTD="Previous.FLUTD.signs.")%>%
  rename(HospUO="Re.UO.in.hosp")%>%
  rename(PrevUO="Previous.UO")%>%
  rename(Flush="Flush.")%>%
  rename(Bolus="Bolus.")%>%
  rename(hrsCath="Hrs.of.cath")%>%
  rename(cathSize="Size.of.u.cath")%>%
  filter(!is.na(cathSize)) %>%
  filter(!is.na(  hrsCath)) %>%
  filter(!is.na( RepeatUO)) %>%
  filter(!is.na( Survival30)) %>%
  filter(!is.na( FLUTD)) %>%
  filter(!is.na( PrevUO)) %>%
  filter(!is.na( Flush)) %>%
  filter(!is.na(Bolus)) %>%
  filter(!is.na( HospUO )) %>%
  select(cathSize,  hrsCath, RepeatUO, Survival30, FLUTD, PrevUO, Flush, Bolus, HospUO ,Ket, Val, Opi, Dex, Iso, Prop)
CatDatasetCreat$RepeatUO <- ifelse(CatDatasetCreat$RepeatUO == "Yes", 1, ifelse(CatDatasetCreat$RepeatUO == "No", 0,2))
CatDatasetCreat$Survival30 <- ifelse(CatDatasetCreat$Survival30 == "Yes", 1, ifelse(CatDatasetCreat$Survival30 == "No", 0,2))
CatDatasetCreat$FLUTD <- ifelse(CatDatasetCreat$FLUTD == "Yes", 1, ifelse(CatDatasetCreat$FLUTD == "No", 0,2))
CatDatasetCreat$PrevUO <- ifelse(CatDatasetCreat$PrevUO == "Yes", 1, ifelse(CatDatasetCreat$PrevUO == "No", 0,2))
CatDatasetCreat$HospUO <- ifelse(CatDatasetCreat$HospUO == "Yes", 1, ifelse(CatDatasetCreat$HospUO == "No", 0,2))
CatDatasetCreat$Flush <- ifelse(CatDatasetCreat$Flush == "Yes", 1, ifelse(CatDatasetCreat$Flush == "No", 0,2))
CatDatasetCreat$cathSize <- ifelse(CatDatasetCreat$cathSize == "Slippery sam", "3.5", ifelse(CatDatasetCreat$cathSize == "3.5", "3.5", 5))
CatDatasetCreat$Bolus <- ifelse(CatDatasetCreat$Bolus== "Yes", 1, ifelse(CatDatasetCreat$Bolus == "No", 0,2))
CatDatasetCreat$cathSize<- as.factor(CatDatasetCreat$cathSize)
CatDatasetCreat
```

```{r echo=FALSE}
cat_data <- cats %>%
  select(Age, Weight..lbs., Breed, Diet, ER.Drugs., Difficulty, Time..min., Size.of.u.cath, Other.cath, Prescription.diet., Medications.after.dsg, Ketamine, Opioid, Isoflurane, Valium, Dexmedetomidine, Propofol, Water.Change, Environment.Change, Litter.Change, Signs.at.home, Repeat.UO., Survival.30.day, Readmit.)


```

### Overview
  Our dataset focuses on cats diagnosed with urinary tract disease, specifically those that have experienced urinary obstructions (UO) requiring hospitalization. While these cats receive treatment, some are returning to the hospital post-procedure, and the underlying reasons for these repeat visits remain unclear. For examining urinary tract infection in cats we used a dataset of 105 male cats with different variables from measurements taken at the doctors office and input from the owners of the cats based on their home life. We wanted to look at if there were any changes doctors could make in terms of care for the cats as well as how the home life of these cats can affect their health. 

### Variables of Interest
After testing the data set and examining possible variables we settled on the relevant variables: Diet, Prescription Diet, Water Change, Environment Change, and Litter Change along with symptoms variables such as Hematuria, Lethargy, Polakiuria, Strainguria. All these variables were used to describe why outcomes such as variables like Repeat Urinary Tract Infection, Readmittance into the Hospital, and the Survival of Cats after 30 days occur. 

### Analysis
  First we wanted to look at vet care of cats and more importantly how diets and prescription diets affected the cats frequency of getting a Urinary Tract infections after the surgery. As show by the charts it appear that cats are less likely to get urinary tract infections when consuming both dry and wet foods. In terms of prescription diets it appears that even though there are any cat getting repeat urinary tract infection that don't have a prescription diet, many cats that get the surgery are precribed to go on this diet and  

```{r echo=FALSE, fig.align='center'}
# filter out NAs
cat_data_diet <- cats %>%
  select(Diet, Repeat.UO.) %>%
  filter(!is.na(Diet), !is.na(Repeat.UO.))

# Visualize the distribution of RepeatUO by Diet
ggplot(cat_data_diet, aes(x = Diet, fill = factor(Repeat.UO.))) +
  geom_bar(position = "fill") + 
  labs(
    title = "Diet vs. Repeat UO",
    x = "Diet",
    y = "Proportion",
    fill = "Repeat UO"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal()

# filter out NAs
cat_data_prescription <- cats %>%
  select(Prescription.diet., Repeat.UO.) %>%
  filter(!is.na(Prescription.diet.), !is.na(Repeat.UO.))

# Visualize the relationship between Prescription Diet and Repeat UO
ggplot(cat_data_prescription , aes(x = Prescription.diet., fill = factor(Repeat.UO.))) +
  geom_bar(position = "fill") +  
  labs(
    title = "Prescription Diet vs. Repeat UO",
    x = "Prescription Diet",
    y = "Proportion",
    fill = "Repeat UO"
  ) +
  scale_y_continuous(labels = scales::percent_format()) + 
  theme_minimal()
```


Next we wanted to look at whether the presence of symptoms such as Hematuria, Lethargy, Pollakiura, Strainguria correlated to whether or not cats have a repeat Urinary Tract Infection. [EXPLAIN DIFFERENT SYMPTOMS AND WHAT THEY ARE]

```{r echo=FALSE, fig.align='center'}
# Reshape data into long format for better visualization
cat_data_urine_symptoms <- cats %>%
  select(Repeat.UO., Strainguria, Hematuria, Lethargy, Pollakiuria) %>%
  pivot_longer(cols = -Repeat.UO., names_to = "Symptom", values_to = "Presence") %>%
  na.omit()

# Stacked bar chart showing proportion of cats with symptoms by Return UO status
ggplot(cat_data_urine_symptoms, aes(x = Symptom, fill = factor(Presence))) +
  geom_bar(position = "fill") +
  facet_wrap(~Repeat.UO., labeller = labeller(Repeat.UO. = c("No" = "Symptom Not Present", "Yes" = "Symptom Present"))) +  # Separate panels for Repeat UO Yes/No
  labs(
    title = "Urinary Symptoms and Repeat UO",
    x = "Urinary Symptoms",
    y = "Proportion of Repeat UO's",
    fill = "Repeat UO"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

As shown by the graph, we can see that if the cats are more likely to develop another urinary tract infection if they are experiencing strainguria (difficulty urinating), hematuria (blood in urine), or lethargy(lack of energy) based on the proportions shown above. We also see that there is no need to be alarmed if the cat has pollakiuria (frequently urinating) because this may be because of an aftermath of the surgery and may have no impact on the future possibility of a UO on the cat because the same proportion of cats with and without the Repeat UO have the symptom of Pollakiuria present.


We also wanted to examine how types of change in the home affected the cat in terms of problems afterwards. We defined problems as whether the cat had one of three variables present - Did Not Survive 30 Days, Readmitted into Hospital, or Had a Repeat Urinary Tract Infection. Based on the charts below we can see that when cats do have one of these problems present around 35% of the time there is usually a change in the litter of cat.  

```{r echo=FALSE, fig.align = 'center'}
catsrepeatproblems <- cat_data %>%
  mutate(Any_Yes = if_else(Repeat.UO. == "Yes" | Survival.30.day == "No" | Readmit. == "Yes", "Yes", "No"))

# Filter data to include only cases where Any_Yes is "Yes"
cat_data_yes <- catsrepeatproblems %>%
  filter(Any_Yes == "Yes") %>%
  select(Environment.Change, Litter.Change, Water.Change) %>%
  pivot_longer(cols = everything(), names_to = "Change_Type", values_to = "Change_Present") %>%
  filter(Change_Present %in% c("Yes", "No"))%>%
  na.omit()%>%
  mutate(Change_Type = recode(Change_Type, 
                              "Environment.Change" = "Environmental Change", 
                              "Litter.Change" = "Litter Change", 
                              "Water.Change" = "Water Change"))

# Create a bar chart for distribution of changes at home when Any_Yes is "Yes"
ggplot(cat_data_yes, aes(x = Change_Type, fill = Change_Present)) +
  geom_bar(position = "fill") +  # Normalize bars to show proportions
  labs(
    title = "Changes at Home vs. Problems are Present",
    x = "Type of Change at Home",
    y = "Proportion of Cats Where Home Problem Present",
    fill = "Problem Present"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +  # Show percentage
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

